`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities

### 2019-11-24 ms.Rmd
update ms [skip ci]

---
title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
author: "Daijiang Li, Russell Dinnage, Lucas Nell, Anthony Ives"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    number_sections: no
    theme: sandstone
    toc: yes
    css: style.css
    keep_md: no
  bookdown::word_document2: 
    toc: no
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
fontsize: 12pt
link-citations: yes
always_allow_html: yes
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/methods-in-ecology-and-evolution.csl
bibliography: ref.bib
---

@@

1. Model-based approaches are increasingly popular in ecological studies. A good example of this trend is the use of joint species distribution models to ask questions about ecological communities. However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition. In part, this is due to lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
2. To fill this gap, the R package `phyr` provides implementations of phylogenetic generalized linear mixed models. It also provides functions to calculate community phylogenetic diversity, correlations among functional traits while accounting for phylogenetic relationships, and generalized models for analyzing phylogenetic comparative data.
3. Other programs can also be used to fit phylogenetic species distribution models, but they require considerable programming and expertise. The model formula syntax we propose serves as a simple and unified framework for phylogenetic species distribution modelling.
4. `phyr` can help current and future researchers formulate and analyze phylogenetic joint species-distribution models.

@@

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized; @ovaskainen2011; @warton2015so]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow. The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. `pglmm()` allows the formulation of a diverses set of phylogenetic generalized linear models that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data. `cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait. We then compare `pglmm()` and `cor_phylo()` to methods and programs that are currently available. Finally, we apply  `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

@@

`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g., `psv()`, `pse()`). The workhorse functions of all groups are written in C++ to increase computational speed. Here, we will focus on the first two groups of functions, because they are more complicated and less readily available to practitioners than community phylogenetic diversity metrics.

@@

```{r mainFunc, echo=FALSE}
read.csv("phyr_func.csv", check.names = FALSE) %>% 
  knitr::kable(caption = "(ref:maunFuncCaption)", booktab = T) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
                latex_options = c("striped")) %>% 
  kableExtra::column_spec(2, width = "25%")
```

@@

`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species. The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`. `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be phylogenetic covariance matrices or any other covariance matrices that the user defines (e.g., spatial or temporal autocorrelation matrix). `pglmm()` can also fit models with “nested” covariance structures (e.g., a species phylogenetic covariance matrix nested within a site covariance matrix). `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, binary, binomial or Poisson, and Bayesian mode with the addition of zero-inflated binomial and Poisson distributions. Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g. via R package `brms` [@brmsR2018]).

@@

With `bayes = FALSE`, `pglmm()` is fitted using a frequentist approach. ML or REML is used for fitting, with `REML = TRUE` as the default. For a non-Gaussian model (e.g., `family = 'binomial'`), an iterated quasi-likelihood method is used for model fitting which gives the approximate likelihood; p-values for the fixed effects are given by a Wald test and for the random effects by profile likelihood, although we recommend bootstrap-based tests when computationally feasible. Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer`) due to the algorithm used. With `bayes = TRUE`, a Bayesian approach is implemented using INLA [@rue2009approximate], which gives parameter estimates and credible intervals. For large problems with the number of species-site combinations exceeding 2000, the Bayesian computations are considerably faster than the frequentist computations. Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects. Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects (Fig. \@ref(fig:designPlot)).

@@

```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
knitr::include_graphics(normalize_path("designPlot2.png"))
```

Whereas `pglmm()` is designed to accept community composition data, in which the same species can occur in multiple sites, the algorithm used by `pglmm()` can equally be used for comparative data in which each species is represented by only a single data point. `pglmm.compare()` is a wrapper for `pglmm()` that is tailored for comparative data and thus provides an easy-to-use function for analyzing non-Gaussian phylogenetic data.

@@

Earlier versions of `pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` [@pearse2015pez] and `ape` [@ape2018], respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility. Both have new syntax that makes them more intuitive to use. `pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig. \@ref(fig:designPlot)), making model interpretation easier. Both are now coded in C++ (for key functions), which speeds computation time by 5-10X. `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA that is particularly useful for large datasets. Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.

@@

`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4`, although it allows the specification of phylogenetic covariance matrices. `pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities. Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present. Nested models are structurally incompatible with the architecture of `lme4`.

@@

### pglmm as a Joint Species Distribution Model (JSDM)

@@

We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites. We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument. Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites. This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.

```{r}
z <- pglmm(
  abund ~ 1 + envi + trait + envi:trait +
    (1 | species__)  + (1 | site__) + 
    (envi | species__) + (trait | site) +
    (1 | species__@site),
  data = dat, 
  cov_ranef = list(species = phy, site = Vspace)
)
summary(z)
```

The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other related statistics (AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties.

@@

Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) for 50 species. We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.

@@

No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/web/packages/phyr/index.html).

>>>>

---
title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
author: "Daijiang Li, Russell Dinnage, Lucas Nell, Matthew Helmus, Anthony Ives"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::word_document2: 
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: sandstone
    toc: yes
    css: style.css
    keep_md: no
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  redoc::redoc:
fontsize: 12pt
link-citations: yes
always_allow_html: yes
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/methods-in-ecology-and-evolution.csl
bibliography: ref.bib
---

@@

1. Model-based approaches are increasingly popular in ecological studies. A good example of this trend is the use of joint species distribution models to ask questions about ecological communities. However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition. In part, this is due to lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
2. To fill this gap, the R package `phyr` (pronounced fire) implements a suite of metrics, comparative methods and mixed models that use phylogenies to understand and predict community composition and other ecological and evolutionary phenomena. The `phyr` workhorse functions are implemented in C++ making all calculations and model estimations fast.
3. `phyr` can fit a variety of models such as phylogenetic joint-species distribution models, spatiotemporal-phylogenetic autocorrelation models, and phylogenetic trait‐based bipartite network models. `phyr` also estimates phylogenetically independent trait correlations with measurement error to test for adaptive syndromes and performs fast calculations of common alpha and beta phylogenetic diversity metrics. All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and the inclusion of phylogenetic terms based on phylogenetic covariance matrices.
4. The functions and model formula syntax we propose in `phyr` serves as a simple and unified framework that ignites the use of phylogenies to address a variety of ecological questions.

@@

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized; @ovaskainen2011; @warton2015so]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow. The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. `pglmm()` allows the formulation of a diverse set of phylogenetic generalized linear models (PGLMM) that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data. `cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait. We then compare `pglmm()` and `cor_phylo()` to methods and programs that are currently available. Finally, we apply  `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

@@

`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic comparative methods (`cor_phylo()` and `pglmm_compare()`), and community phylogenetic diversity metrics (e.g., `psv()`, `pse()`). The workhorse functions of all groups are written in C++ to increase computational speed. Here, we will focus on the first two groups of functions (especially `pglmm()` and `cor_phylo()`), because they are more complicated and less readily available to practitioners than community phylogenetic diversity metrics.

@@

```{r mainFunc, echo=FALSE}
read.csv("phyr_func.csv", check.names = FALSE) %>% 
  knitr::kable(caption = "(ref:maunFuncCaption)") #%>%
  # kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
  #               latex_options = c("striped")) %>% 
  # kableExtra::column_spec(2, width = "25%")
```

@@

`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species. The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`. `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be phylogenetic covariance matrices or any other covariance matrices that the user defines (e.g., spatial or temporal autocorrelation matrix). `pglmm()` can also fit models with “nested” covariance structures (e.g., a species phylogenetic covariance matrix nested within a site covariance matrix). `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, binary, binomial or Poisson, and Bayesian mode with the addition of zero-inflated binomial and Poisson distributions. Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan [e.g. via R package `brms` @brmsR2018].

@@

With `bayes = FALSE`, `pglmm()` is fitted using a frequentist approach. ML or REML is used for fitting, with `REML = TRUE` as the default. For a non-Gaussian model (e.g., `family = 'binomial'`), an iterated quasi-likelihood method is used for model fitting which gives the approximate likelihood; p-values for the fixed effects are given by a Wald test and for the random effects by profile likelihood, although we recommend bootstrap-based tests when computationally feasible. Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer`) due to the algorithm used. With `bayes = TRUE`, a Bayesian approach is implemented using INLA [@rue2009approximate], which gives parameter estimates and credible intervals. For large problems with the number of species-site combinations exceeding 2000, the Bayesian computations are considerably faster than the frequentist computations. Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects. Therefore, `pglmm()` has associated plotting functions `pglmm_plot_ranef()` that present the design matrices for the random effects (Fig. \@ref(fig:designPlot)).

@@

```{r designPlot, echo = F, fig.cap="(ref:designPlotCap)", out.width="100%"}
knitr::include_graphics(normalize_path("designPlot2.png"))
```

Whereas `pglmm()` is designed to accept community composition data, in which the same species can occur in multiple sites, the algorithm used by `pglmm()` can equally be used for comparative data in which each species is represented by only a single data point. `pglmm_compare()` is a wrapper for `pglmm()` that is tailored for comparative data and thus provides an easy-to-use function for analyzing non-Gaussian phylogenetic data.

@@

Earlier versions of `pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` [@pearse2015pez] and `ape` [@ape2018], respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility. Both have new syntax that makes them more intuitive to use. `pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig. \@ref(fig:designPlot)), making model interpretation easier. Both are now coded in C++ (for key functions), which speeds computation time by 5-10X. `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA [@rue2009approximate] that is particularly useful for large datasets. Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.

@@

`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4` [@bates2015lme4], although it allows the specification of phylogenetic covariance matrices. `pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities. Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present. Nested models are structurally incompatible with the architecture of `lme4`.

@@

### PGLMM as a Joint Species Distribution Model (JSDM)

@@

We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites. We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument. Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites. This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.

```{r, eval=FALSE}
z <- pglmm(
  abund ~ 1 + envi + trait + envi:trait +
    (1 | species__)  + (1 | site__) + 
    (envi | species__) + (trait | site) +
    (1 | species__@site),
  data = dat, 
  cov_ranef = list(species = phy, site = Vspace)
)
```

```{r, echo=FALSE}
z = readRDS("z.rds")
```

```{r}
summary(z)
```

The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other related statistics (AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties. These results show that `pglmm()` correctly recovered that the hypothetic functional trait interacted with environmental variable to affect species composition.

@@

Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) for 50 species. We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively. We also set their measurement errors to be 0.2 and 1, respectively.

@@

No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/package=phyr).
